Relationships in the extremes and their influence on competition and coexistence

### 2021-07-14 paper/Paper.Rnw
working on sentence structure in intro

The maintenance of biodiversity in competitive ecosystems is a central idea in ecology. Among others, the modern theory of coexistence seeks to assess and quantify the mechanisms that result in coexisting species and thus the maintenance of biodiversity. 
Many are familiar with the principle of competitive exclusion - an idea which postulates if species within the same trophic level and similar niches exist, one will eventually outcompete the other (Hardin 1960). Modern coexistence theory provides a quantitative breakdown of contributing mechanisms as criteria for coexistence and competitive exclusion (Chesson 1994).

Equally central to ecology is the study of relationships between ecological variables; however, it is almost exclusively blanketed by simple measures of associativity. Recent work has shown that tail associations, or associations between the tail ends of data, and specifically the asymmetry between tail associations have important considerations for ecological phenomena; although, to our knowledge, the consequences/implications of asymmetric tail associations for competition and coexistence have not yet been investigated. 
Correlation measures such as Pearson or Spearman correlation disguise the asymmetry of tail associations, and datasets with different tail associations can have the same correlation coefficient (see Figure). Asymmetry in tail associations describes tighter correlations in regions of a dataset compared to other regions of the dataset. A simple example of asymmetric tail associations is a dataset which is more correlated in the lower (upper) half than the upper (lower) half, described as left (right) tail associated. Current research has shown that ATAs have important considerations for extinction risks, “skewness of spatial averages”, Taylor’s Law, Moran Effect/ spatial synchrony (Gosh et al 2020 Ecosphere, Gosh et al 2020 And. EcoRes,). Despite the existence of tail associations within ecological data, to our knowledge, the implications of tail associations for competition and coexistence has not yet been examined; doing so is the main goal of this paper.

>>>>

Central to ecology is the maintenance of biodiversity in competitive ecosystems. 
Populations subject to strong competitive pressure may succumb to competitive exclusion - an idea which postulates if species within the same trophic level and similar niches exist, one will eventually outcompete the other (Hardin 1960).
Mediation of competitive pressure and exclusion allow for species to stably coexist. 
The criteria for coexistence and competitive exclusion can be broken down into contributing mechanisms provided by modern coexistence theory, a framework which seeks to assess and quantify mechanisms that result in coexisting species and thus the maintenance of biodiversity (Chesson 1994).

Equally central to ecology is the study of relationships between ecological variables; however, it is almost exclusively blanketed by simple measures of associativity, i.e., correlation coefficients.
Although two sets of variables may have equal correlation coefficients, they can show different association structures (see Figure).
More specifically, the association within the tail ends of data and the asymmetry between the tail associations, termed asymmetric tail associations (ATAs), tell a more interesting story about the relationship between variables...and?...
Recent work has shown that ATAs have important considerations for ecological phenomena; although, the implications of ATAs for competition and coexistence have not yet been investigated. 
Asymmetry of tail associations within a dataset of variables describes uneven correlations between the tails of a dataset.
For a simple example, a dataset with more association in the lower (upper) half and lesser association in the upper (lower) half is termed left (right) tail associated. 
The application of ATAs in ecology is relatively new, and in exploration of ATAs in ecology, Gosh et al (AER) found ATAs in all eight ecological datasets they considered.
Thus, the existence and functioning role of ATAs in nature is reasonable to assume and question. 
In fact, ATAs can have influence on ecosystem service stability, Taylor’s law, and extinction risks via extreme climatic events. Conversely, ATA’s can be influenced by Moran Effect and Leibeig’s Law of Minimum (Gosh et al 2020 Ecosphere, Gosh et al 2020 And. EcoRes,).
Despite the frequent occurrence of ATAs in ecological data, the implications of tail associations for competition and coexistence, to our knowledge, has not yet been examined; doing so is the main goal of this paper.
